Follow-up Samples

### 2017-03-23 submission/manuscript_outline_20161024.Rmd

# Main CRC bacteria differences between Adenoma and Carcinoma
main_crc_bugs_summary <- read.csv('../data/process/tables/adn_crc_maybe_pvalue_summary.csv', header = T, row.names = 1)
crc_bug_info <- read.csv("../data/process/tables/adn_crc_maybe_diff.csv", header = T, stringsAsFactors = F)

# Lesion Model 
lesion_top10_varsMDA <- read.csv('../data/process/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T)
lesion_imp_vars <- read.csv('../data/process/tables/rf_otu_tax.csv', header = T, row.names = 1)
lesion_AUC_run_summary <- read.csv('../data/process/tables/Reduced_ROC_model_summary.csv', header = T)

# IF Model
IF_top10_varsMDA <- read.csv('../data/process/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T)
IF_imp_vars <- read.csv('../data/process/tables/if_rf_otu_tax.csv', header = T, row.names = 1)
IF_AUC_run_summary <- read.csv('../data/process/tables/reduced_IF_ROC_model_summary.csv', header = T)

# All model information
all_probs_paired_wilcox <- read.csv('../data/process/tables/all_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
all_pred_stat_summary <- read.csv('../data/process/tables/all_models_summary_info.csv', header = T, stringsAsFactors = F)
all_confu_matrix <- read.csv('../data/process/tables/all_models_confusion_summary.csv', header = T, stringsAsFactors = F)

# Common Variable Info
common_vars_summary <- read.csv('../data/process/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)

#Therapy Info
chemo_rad_summary <- read.csv('../data/process/tables/probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
chemo_rad_common_vars_summary <- read.csv('../data/process/tables/chemo_rads_treatment_pvalue_summary.csv', header = T, stringsAsFactors = F)

@@

**Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.  This study tested the hypothesis that treatment for adenoma or carcinoma result in changes to the bacterial community.  Specifically, we tried to identify components within the community that were different before and after removal of lesion (adenoma or carcinoma).

**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma cases (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.05, paste("<"), paste(">"))` 0.05).  Yet no difference was found in the relative abundance of any OTU before and after treatment for adenoma or carcinoma groups (P-value `r ifelse(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']) < 0.05, paste("<"), paste(">"))` 0.05).  A lesion model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` and follow up samples had no difference in the positive probability of lesion versus initial samples (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that the lesion associated community persists after treatment.  A treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` - `r format(max(IF_AUC_run_summary$ROC), digits = 3)` and had a decrease positive probability for the follow up samples to be an initial sample (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05); suggesting that there are members within the community that repond to treatment.  The lesion model used a total of `r length(lesion_top10_varsMDA$variable)` variables while the initial sample model used a total of `r length(IF_top10_varsMDA$variable)` variables.  A total of `r length(common_vars_summary$otu)` OTUs were common to both models with many classifying to commensal bacteria (e.g. *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*).

**Conclusions:** Our data partially supports the hypothesis that the bacterial community changes after treatment.  Individuals with carcinoma have more drastic differences to the overall community then those with adenoma.  Commensal bacteria were crucial for accurate model classification, suggesting that these bacteria may be important to initial polyp formation and transition to carcinoma.

@@

Recent studies in humans have shown that both the bacterial community and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].  Further, Dejea, et al. observed that bacterial communities are altered between normal and tumor tissue [@dejea_microbiota_2014].  Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.  Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest that biomarkers using our microbes could be developed.  Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies show how our gut bacteria can impact CRC progression via a changed community or invasion by more inflammatory bacteria [@flynn_metabolic_2016].  They provide very little information as to whether these changed communities rebound after successful treatment of lesion (adenoma or carcinoma).

In this study, we tested the hypothesis that there are detectable changes to the bacterial community between pre- (initial) and post- (follow up) treatment of lesion.  We analyzed changes in alpha and beta diversity as well as the relative abundance of specific Operational Taxonomic Units (OTUs).  We then investigated several important genera from oral microbes that have been suggested to be important in CRC pathogenesis [@flynn_metabolic_2016].  We next utilized Random Forest to build two models: The first was built to classify lesion versus non-lesion (normal) while the second was built to classify initial versus follow up treatment samples.  Subsequent observations on how these models and OTUs within them performed before and after treatment helped inform us as to whether they were changing after treatment and whether it was towards a normal community.  We also investigated the two models for similar important OTUs to identify which were important for classifying both lesion versus normal and initial versus follow up treatment samples.  Finally, since treatment also included chemotherapy and radiation therapy along with lesion removal, we tested if these additionally treatments significantly lowered any of the metrics examined versus those who received only lesion removal.  This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists.

@@

***The Bacterial Community:***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma (n = `r length(rownames(filter(good_metaf, Diagnosis == "adenoma")))`) or carcinoma (n = `r length(rownames(filter(good_metaf, Diagnosis != "adenoma")))`) had any broad differences between their initial and follow up samples.  We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma (P-value `r ifelse(change_summary['thetayc', 'Pvalue'] < 0.001, paste("<"), paste(">"))` 0.001) [Figure 1A].  The bacterial community structure before and after surgery was visualized using NMDS for both adenoma [Figure 1B] (PERMANOVA `r ifelse(beta_summary['Adenoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05) and carcinoma [Figure 1C] (PERMANOVA `r ifelse(beta_summary['Carcinoma', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity (PERMANOVA `r ifelse(beta_summary['All', 'PERMANOVA'] < 0.05, paste("<"), paste(">"))` 0.05).  There was no difference between initial and follow up samples when investigating alpha diversity metrics for lesion, adenoma only, or carcinoma only for any metric tested [Table S1].  Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for lesion, adenoma only, or carcinoma only [Figure S1].

***Carcinoma Associated Genera:*** 
For carcinoma [Figure S2A] (P-value `r ifelse(main_crc_bugs_summary["parv", "crc_BH"] < 0.05 & main_crc_bugs_summary["porp", "crc_BH"] < 0.05 & main_crc_bugs_summary["pept", "crc_BH"] < 0.05, paste("<"), paste(">"))` 0.05) but not adenoma [Figure S2B] (P-value `r ifelse(main_crc_bugs_summary["parv", "adn_BH"] < 0.05 & main_crc_bugs_summary["porp", "adn_BH"] < 0.05 & main_crc_bugs_summary["pept", "adn_BH"] < 0.05, paste("<"), paste(">"))` 0.05) there was a significant decrease in *Porphyromonas*, *Parvimonas*, and *Peptostreptococcus* between initial and follow up samples [Figure S2].  *Fusobacterium* for both adenoma and carcinoma had no significant differences between initial and follow up samples [FIgure S2] (P-value `r ifelse(main_crc_bugs_summary["fn", "crc_BH"] < 0.05 & main_crc_bugs_summary["fn", "adn_BH"] < 0.05, paste("<"), paste(">"))` 0.05).  Although there were significant differences, only a small percentage of those with adenoma or carcinoma were positive for any of these specific genera [Figure S2].

***The Lesion Model:***
We tested whether follow up samples were more normal than initial samples by building a model based solely on OTUs alone that classified lesion versus normal.  This model had an AUC range of `r format(min(lesion_AUC_run_summary$ROC), digits = 3)` - `r format(max(lesion_AUC_run_summary$ROC), digits = 3)` after 100 iterations of 20 repeated 10-fold cross validations.  The ROC curve for the final lesion model used was within the observed range of the 100-different test set AUC iterations [Figure 2A].  There was a total of `r length(lesion_top10_varsMDA$variable)` OTUs that were used in this model [Figure 2B].  *`r gsub("_unclassified", "", lesion_imp_vars[as.character((filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(lesion_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) resulted in the largest decrease in MDA [Figure 2B].  Heavily represented genera within the model included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.

If there was movement towards a more normal community after treatment of lesion, we would expect to find a decrease in the positive probability of the follow up sample to be a lesion.  We observed no such decrease in positive probability for either those with adenoma or carcinoma [Figure 2C & 2D] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "carcinoma_only")[, "BH_correction"] < 0.05 & filter(all_probs_paired_wilcox, model_type =="red_lesion" & comparison == "all_adenoma")[, "BH_correction"] < 0.05, paste("<"), paste(">"))` 0.05).  We also observed a significant difference between the predicted and actual calls (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_lesion" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).  However, this model correctly classified the one individual who still had carcinoma on follow up and this individual's positive probability of lesion increased between their initial and follow up sample [Figure 2C].

***The Treatment Model:***
To test if there were differences that could be used to classify initial and follow up treatment samples we built a model to classify initial samples.  This treatment model had an AUC range of `r format(min(IF_AUC_run_summary$ROC), digits = 3)` to `r format(max(IF_AUC_run_summary$ROC), digits = 3)`.  after 100 iterations of 20 repeated 10-fold cross validations.  The test set AUC range for this model performed better than the training set AUCs.  There was a marked decrease in the ROC curve for the final model used when compared to the 100 test set AUC iterations [Figure 3A].  There was a total of `r length(IF_top10_varsMDA$variable)` OTUs that were used for this model [Figure 3B].  The variable that resulted in the largest MDA was *`r gsub("_unclassified", "", IF_imp_vars[as.character((filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable))[, 1]), "Genus"])`* (`r filter(IF_top10_varsMDA, variable != "fit_result") %>% top_n(1, mean_MDA) %>% select(variable)`) [Figure 3B].  Similar to the lesion model, heavily represented genera included OTUs from *Lachnospiraceae*, *Bacteroides*, *Roseburia*, *Blautia*, and *Ruminococcus*.

If there were changes between initial and follow up samples we would expect the positive probability of being an initial sample to be decreased in the follow up.  The is what we observed for the treatment model (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "lesion")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  When we separated lesion into adenoma and carcinoma there was a decrease in positive probability for both the carcinoma [Figure 3C] (P-value `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "carcinoma_only")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001) and adenoma group [Figure 3D] `r ifelse(filter(all_probs_paired_wilcox, model_type =="red_IF" & comparison == "all_adenoma")[, "BH_correction"] < 0.001, paste("<"), paste(">"))` 0.001).  For this model, there was no difference between the predicted and actual classifications (P-value `r ifelse(filter(all_pred_stat_summary, samples_tested == "all" & model_type == "red_IF" & info == "McnemarPValue")[, "all"] < 0.05, paste("<"), paste(">"))` 0.05).

***Common OTUs to both Models:***
We next wanted to know what predictors within the lesion model were also in the initial sample model.  The main purpose was to identify which OTUs could be important in lesion formation and were impacted by treatment.  When we compared the two different models with each other there were a total of `r length(common_vars_summary$otu)` common OTUs.  Some of the most common taxonomic identifications belonged to *`r filter(common_vars_summary, otu == "Otu000028")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000061")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000037")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000006")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000015")[, "lowest_ID"]`*, *`r filter(common_vars_summary, otu == "Otu000167")[, "lowest_ID"]`*, and *`r filter(common_vars_summary, otu == "Otu000005")[, "lowest_ID"]`*.  These along withthe vast majority of the OTUs that were common between models had classifications to bacteria typically thought of as commensal [Table S2].

***Chemotherapy and Radiation Differences:*** 
After observing these changes due to treatment we assessed whether chemotherapy or radiation impacted the observed results.  Only the treatment model had a significant change in positive probability for those treated with chemotherapy or radiation versus those who received neither [Figure 4A & Table S3] (P-value `r ifelse(chemo_rad_summary['chemo_IF', 'bh'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05). Additionally, there was no difference between the change in positive probability for chemotherapy and radiation therapy (Uncorrected P-value `r ifelse(chemo_rad_summary['chemo_v_rads_IF', 'pvalue'] < 0.05 & chemo_rad_summary['rad_IF', 'bh'] < 0.05, paste("<"), paste(">"))` 0.05).  Using the `r length(common_vars_summary$otu)` common OTUs we tested how chemotherapy or radiation affects important members of the community.  For both chemotherapy and radiation, the directionality of change of these OTUs were similar [Figure 4B & 4C].  After multiple comparison correction, only *`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "lowest_ID"]`* (`r filter(chemo_rad_common_vars_summary, bh_chemo < 0.05)[, "otu"]`) remained significantly different between those who did and did not receive chemotherapy.  This data suggests that follow up samples from those treated with either chemotherapy or radiation may have had a larger change from initial sample due to effects on the bacterial community.

>>>>

# Adenoma Model 
adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
adn_imp_vars <- read.csv('../data/process/tables/adn_rf_otu_tax.csv', header = T, row.names = 1)
adn_AUC_run_summary <- read.csv('../data/process/tables/adn_Reduced_ROC_model_summary.csv', header = T)

# SRN Model
srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T)
srn_imp_vars <- read.csv('../data/process/tables/srn_rf_otu_tax.csv', header = T, row.names = 1)
srn_AUC_run_summary <- read.csv('../data/process/tables/srn_Reduced_ROC_model_summary.csv', header = T)

# CRC Model
crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T)
crc_imp_vars <- read.csv('../data/process/tables/crc_rf_otu_tax.csv', header = T, row.names = 1)
crc_AUC_run_summary <- read.csv('../data/process/tables/crc_Reduced_ROC_model_summary.csv', header = T)

# All model information
all_probs_paired_wilcox <- read.csv('../data/process/tables/all_crc_srn_adn_models_wilcox_paired_pvalue_summary.csv', header = T, stringsAsFactors = F)
all_pred_stat_summary <- read.csv('../data/process/tables/all_crc_srn_adn_models_summary_info.csv', header = T, stringsAsFactors = F)
all_confu_matrix <- read.csv('../data/process/tables/all_crc_srn_adn_models_confusion_summary.csv', header = T, stringsAsFactors = F)

# Common Variable Info
common_vars_summary <- read.csv('../data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)

#Therapy Info
chemo_rad_summary <- read.csv('../data/process/tables/crc_probs_chemo_rad_pvalue_summary.csv', header = T, row.names = 1)
chemo_rad_common_vars_summary <- read.csv('../data/process/tables/chemo_rads_treatment_pvalue_summary.csv', header = T, stringsAsFactors = F)

comb_adn_surg_summary <- read.csv('../data/process/tables/adn_combined_probs_surgery_pvalue_summary.csv', header = T, stringsAsFactors = F, row.names = 1)

comb_adn_surg_common_vars_summary <- read.csv('../data/process/tables/all_adn_surg_pvalue_summary.csv', header = T, stringsAsFactors = F)

@@

**Background:** Colorectal cancer (CRC) is a worldwide health problem and research suggests a correlation between the fecal bacterial microbiome and CRC.  This study tested the hypothesis that treatment for adenoma or carcinoma results in changes to the bacterial community.  Specifically, we tried to identify components within the community that were different before and after treatment of adenoma, advanced adenoma (Screen Relevant Neoplasia (SRN)), and carcinoma.

**Results:** There was a larger change in the bacterial community in response to treatment for carcinoma versus adenoma or SRN cases *(P-value )*.  But there was a trend for increasingly less community similarity between initial and follow up samples from adenoma to SRN to carcinoma.  Despite this, no difference was found in the relative abundance of any OTU before and after treatment for adenoma, SRN, or carcinoma groups *(P-value )*.  Using Random Forest models to assess whether changes in follow up samples were towards a normal community showed that only those with carcinoma had a significant decrease in positive probability *(P-value)*; providing further evidence that treatment has the greatest effect in those with carcinoma. The adenoma model used a total of *model n* OTUs, the SRN model used a total of *model n* OTUs, and the carcinoma model used a total of *model n* OTUs.  A total of *common n* OTUs were common to both models with many classifying to commensal bacteria (e.g. *Lachnospiraceae*, *Bacteroides*, *Anaerostipes*, *Blautia*, and *Dorea*).  Both chemotherapy and radiation did not provide any additional changes to the bacterial community in those treated for carcinoma *(Pvalue )*

**Conclusions:** Our data partially supports the hypothesis that the bacterial community changes after treatment.  Individuals with carcinoma had more drastic differences to the overall community then those with adenoma.  Common OTUs to all models were overwhelmingly from commensal bacteria, suggesting that these bacteria may be important to initial polyp formation, development of advanced adenoma, and transition to carcinoma.

@@

Recent studies in humans have shown that both the microbiome and specific members within it correlate with CRC pathogenesis [@zackular_manipulation_2016; @arthur_microbial_2014].  Further, bacterial communities have been observed to be altered between normal and tumor tissue [@dejea_microbiota_2014].  Mouse models of CRC have further demonstrated the importance of the microbiome, both on a community [@zackular_gut_2013; @zackular_manipulation_2016] and species level [@arthur_microbial_2014], for tumorgenesis.  Collectively, these studies provide a tantalizing link between our gut bacteria and CRC and suggest that biomarkers using our microbes could be developed.  Indeed, builidng models using 16S rRNA gene sequencing along with clinical tests such as Fecal Immunoglobulin Test (FIT) result in good predictions of CRC [@baxter_microbiota-based_2016; @zeller_potential_2014].  Although these studies show how our gut bacteria can impact CRC progression via a changed community or invasion by more inflammatory bacteria [@flynn_metabolic_2016].  They provide very little information as to whether these communities change and rebound towards normal after successful treatment of adenoma, advanced adenoma (Screen Relevent Neoplasia (SRN)), or carcinoma.

Providing an answer to this question is important because it has far reaching implications on both how the bacterial community causes the formation of more polyps [@zackular_manipulation_2016; @arthur_microbial_2014] and the ability to be able to use the microbiome as a predictive screening tool [@baxter_microbiota-based_2016; @zeller_potential_2014].  Understanding polyp formation and transition to SRN and then carinoma is crucial to being able to understand how to prevent CRC occurence.  Response of the community to treatment is also eqaully important to predictive models designed for screening purposes since an unresponsive community would provide little additional information for important events, such as recurrence [@hassan_efficacy_2016].

Using pre- (initial) and post- (follow up) treatment samples we tested the hypothesis that treatment causes detectable changes to the microbiome in those with adenoma, SRN, and carincoma.  First, we assessed differences between initial and follow up samples in adenoma, SRN, or carinoma using alpha or beta diversity metrics.  Second, we explored whether models built to classify adenoma, SRN, or carcinoma versus normal were able to identify specific community members that differed between initial and follow up.  We also used these models to assess whether changes in the community were toward a more normal micorbiome.  Finally, we assessed both whether surgery for adenomas and SRN provided larger community changes or whether chemotherapy or radiation provided additive changes to the microbiome over surgerical resection. This study helps to provide evidence as to whether treatment can influence the community and if the CRC microbiome, identified in previous studies, persists after such interventions.

@@

***The Bacterial Community:*** 
Within our `r length(rownames(good_metaf))`-person cohort we tested whether those with adenoma *(n = )*, SRN *(n = )*, or carcinoma *(n = )* had any broad differences between their initial and follow up samples.  We found that those with carcinoma had a more dissimilar bacterial community between their initial and follow up sample than those with adenoma *(P-value 0.001)* [Figure 1A].  Although no significant differences were observed between SRN and carcinoma there was an increase in the dissimilarity of the initial and follow up samples from adenoma *(mean and SD)* to SRN *(mean and SD)* to carcinoma *(mean and SD)* [Figure 1A].  The bacterial community structure before and after surgery was visualized using NMDS for adenoma [Figure 1B] *(PERMANOVA  0.05)*, SRN [FIgure 1C] *(PERMANOVA 0.05)*, and carcinoma [Figure 1C] *(PERMANOVA  0.05)*.  Interestingly, when initial and follow up samples were compared, regardless of whether the lesions were adenoma or carcinoma, there was no significant overall difference in beta diversity *(PERMANOVA  0.05)*.  There was no difference between initial and follow up samples when investigating alpha diversity metrics for adenoma, SRN, or carcinoma for any metric tested [Table S1].  Additionally, there was also no difference in the relative abundance of any OTU between initial and follow up samples for adenoma, SRN, or carcinoma only [Figure S1].

***Adenoma Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2A].  There was a total of *blank* OTUs in this model with the vast majority classifying to bacteria typically thought of as commensal [Figure S3A]. There was a significant difference between the actual and model predicted group stratification calls *(P-value)*.  There was no significant decrease in the positive probability of adenoma between initial and follow up samples *(P_value)* [Figure 2A].

***SRN Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2B].  There was a total of *blank* OTUs in the SRN model [Figure S3B].  Similar to the adenoma model the vast majority of OTUs classified to bacteria typically thought of as commensal. Also similar to the adenoma model there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.  There was no significant decrease in the positive probability of SRN between initial and follow up samples *(P_value)* [Figure 2B].

***Carcinoma Model***
The range of model AUC's from 100 runs of 20 repeated 10 fold cross-validation was *from to this* with the AUC of the model used being *AUC value* [Figure S2C].  Interestingly, the AUCs improved from adenoma to SRN to carcinoma [Figure 2].  There was a total of *blank* OTUs in the carinoma model [Figure S3C].  Similar to the adenoma and SRN models the vast majority of OTUs classified to bacteria typically thought of as commensal but OTUs that also classified to *Fusobacterium*, *Porphyromonas*, and *Parvimonas* appear to be important for carcinoma classification [Figure S3C].  Also similar to the adenoma and SRN models there was a significant difference between the actual and model predicted group stratification calls *(P-value)*.  There was a significant decrease in the positive probability of carcinoma between initial and follow up samples *(P_value)* [Figure 2C]; suggesting that the carcinoma samples changed towards normal after treatment, unlike either adenoma or SRN.

***Adenoma, SRN, and Carcinoma Common OTUs***
We next wanted to know what predictors within the adenoma, SRN, and carcinoma models were similar to each other.  The main purpose was to identify which OTUs could be important at all three stages of disease.  When we compared the three different models with each other there were a total of *number* common OTUs.  Some of the most common taxonomic identifications belonged to *one*, *two*, *three*, *four*, *five*, *six*, and *seven*.  These along with the vast majority of the OTUs that were common between these models had classifications to bacteria typically thought of as commensal [Table S2].

***Treatment Affects on Community***
After observing these these changes from treatment we assessed whether chemotherapy or radiation, in the carcinoma group, and surgery, in the adenoma group, impacted the observed results.  In the carcinoma group neither chemotherapy or radiation provided any additive change from initial sample over surgery alone [Table S3]. For the adenoma group there was no difference between those that received surgery and those that did not [Table S4]. For surgery adenoma and SRN were combined due to the low number of surgery occurances.  There was no difference in the proportion of those receiving surgery between the adenoma and SRN groups *(P-value)*.  This data suggests that microbiome changes observed in the carcinoma group were mostly a result from surgery and not from chemotherapy or radiation.

#Need to make a table with the adenoma results

### 2017-03-23 submission/supplemental_outline_20161024.Rmd

**Table S2: Summary of OTUs in both Lesion and Initial/Follow Up Models**

@@

common_vars_summary <- read.csv('../data/process/tables/pvalue_IF_lesion_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)

lesion_top10_varsMDA <- read.csv('../data/process/tables/reduced_lesion_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)
IF_top10_varsMDA <- read.csv('../data/process/tables/reduced_IF_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)

@@

# Create MDA vectors
lesion_MDA <- filter(lesion_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)

initial_MDA <- filter(IF_top10_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)

st2 <- select(common_vars_summary, otu) %>% 
  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
  mutate(lesion_MDA = lesion_MDA$mean_MDA) %>% 
  mutate(initial_MDA = initial_MDA$mean_MDA) %>% arrange(desc(lesion_MDA))

colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Lesion Model MDA", "Initial Model MDA")

>>>>

**Table S2: Summary of OTUs in Adenoma, SRN, and Carcinoma Models**

@@

common_vars_summary <- read.csv('../data/process/tables/pvalue_adn_srn_crc_common_imp_vars.csv', header = T, row.names = 1, stringsAsFactors = F)

adn_varsMDA <- read.csv('../data/process/tables/adn_reduced_crc_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)

srn_varsMDA <- read.csv('../data/process/tables/srn_reduced_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)

crc_varsMDA <- read.csv('../data/process/tables/reduced_crc_model_top_vars_MDA_Summary.csv', header = T, stringsAsFactors = F)

@@

# Create MDA vectors
adn_MDA <- filter(adn_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
srn_MDA <- filter(srn_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)
crc_MDA <- filter(crc_varsMDA, variable %in% common_vars) %>% slice(match(common_vars, variable)) %>% select(mean_MDA)

st2 <- select(common_vars_summary, otu) %>% 
  mutate(Tax = gsub("_unclassified", "", common_vars_summary$lowest_ID)) %>% 
  mutate(adn_MDA = adn_MDA$mean_MDA, srn_MDA = srn_MDA$mean_MDA, crc_MDA = crc_MDA$mean_MDA) %>% 
  arrange(desc(adn_MDA), desc(srn_MDA), desc(crc_MDA))

colnames(st2) <- c("OTU", "Lowest Taxonomic ID", "Adenoma Model MDA", "SRN Model MDA", "Carcinoma Model MDA")
